---
title: "Software Insteallation"
author: ""
subtitle: ""
format: 
  html:
    mainfont: "12"
  PrettyPDF-pdf:
    keep-tex:  true
    number-sections: true
editor_options: 
  chunk_output_type: console
---


Please use the instructions below to install and check your installation. 

If you run into issues, you can post a question with information about what you tried and what didn't work, on the [course github discussion page]() and we will try reply.

Since 29 April 2025, the latest INLA package is built for R 4.5, so if you're able to upgrade your R installation, please do so to avoid unnecessary issues. The package will in many cases also work with older R versions, but compatibility is sometimes difficult.

## Installing `INLA` and `inlabru`

Due to the work involved in building the binaries for the INLA package C software for different architectures, the INLA package is not on CRAN, but it can be installed from its own distribution
repository. 

1.  Check your `R` version.

2.  Install R-INLA, instructions can be found  [here](https://www.r-inla.org/download-install)

3.  Install `inlabru` (available from CRAN)

```{r}
#| eval: false
# Enable universe(s) by inlabru-org
options(repos = c(
  inlabruorg = "https://inlabru-org.r-universe.dev",
  INLA = "https://inla.r-inla-download.org/R/testing",
  CRAN = "https://cloud.r-project.org"
))

# Install some packages
install.packages("inlabru")
```

3.  Make sure you have the latest `R-INLA`, `inlabru` and `R` versions installed.

4.  Install the following libraries:

## Installation check

Please check your installation using the basic model runs below. If you run into issues, you can post a question with information about what you tried and what didn't work, on the [course github discussion page](https://github.com/inlabru-org/inlabruCourseMay2025/discussions) and Finn will reply when he's able.
Since 29 April 2025, the latest INLA package is built for R 4.5, so if you're able to upgrade your R installation, please do so to avoid unnecessary issues. The package will in many cases also work with older R versions, but compatibility is sometimes difficult.

You can check that `INLA` is correctly installed by running
```{r}
#| eval: false
df <- data.frame(y = rnorm(100) + 10)
fit <- INLA::inla(
  y ~ 1,
  data = df
)
summary(fit)
```
If the simple `inla()` call fails with a crash, you may need to install different
`inla` binaries for your hardware/software combination,
with `INLA::inla.binary.install()`.

When `inla()` works, you can check that `inlabru` is installed correctly by running
the same model in `inlabru`:
```{r}
#| eval: false
fit <- inlabru::bru(
  y ~ Intercept(1, prec.linear = exp(-7)),
  data = df
)
summary(fit)
```