---
title: "Practical 1 - Linear (Mixed) Models"
execute: 
  freeze: auto
format: 
  html:
    mainfont: "12"
  PrettyPDF-pdf:
    keep-tex: false
    number-sections: true
embed-resources: true
editor_options: 
  chunk_output_type: console
---

```{r}
#| echo: false
#| message: false
#| warning: false
#| purl: false
#| results: hide
library(webexercises)

fname <- knitr::current_input()

# Only run purl during render
if (!is.null(fname)) {
  base  <- tools::file_path_sans_ext(basename(fname))
  rfile <- paste0(base, ".R")
  knitr::purl(fname, output = rfile)
}


```


In this practical we are going to fit linear (mixed) models in `inlabru`. 
We are going to to:

- Fit a [simple linear regression](#SLR)
- Fit a [linear regression with discrete covariates and interactions](#LM_int)
- Fit a [linear mixed model](#LMM).



::: {.callout-note icon="false"}
You can download the R-script of this practical by clicking the button below:

{{< downloadthis `r rfile` dname=`r rfile` label="Download R script" >}}
:::



Start by loading useful libraries:

```{r}
#| warning: false
#| message: false
#| code-summary: "Load libraries"

library(dplyr)
library(INLA)
library(ggplot2)
library(patchwork)
library(inlabru)     
# load some libraries to generate nice plots
library(scico)
```


## Simple linear regression{#SLR}

We consider a simple linear regression model with Gaussian observations 

$$
y_i\sim\mathcal{N}(\mu_i, \sigma^2), \qquad i = 1,\dots,N
$$

where $\sigma^2$ is the observation error, and the mean parameter $\mu_i$ is linked to the **linear predictor** ($\eta_i$) through an identity function: 
$$
\eta_i = \mu_i = \beta_0 + \beta_1 x_i.
$$ 

Here $\mathbf{x}= (x_1,\dots,x_N)$ is a continuous covariate and $\beta_0, \beta_1$ are parameters to be estimated.

To finalize the Bayesian model we assign prior distribution as $\tau = 1/\sigma^2\sim\text{Gamma}(a,b)$  and $\beta_0,\beta_1\sim\mathcal{N}(0,1/\tau_{\beta})$ (we will use the default prior settings in INLA for now).

::: {.callout-tip icon="false"}
## {{< bi question-octagon color=#6dc83c >}} Question

What is the dimension of the hyperparameter vector and latent Gaussian field?

`r hide("Answer")`

The hyperparameter vector has dimension 1, $\pmb{\theta} = (\tau)$ while the latent Gaussian field $\pmb{u} = (\beta_0, \beta_1)$ has dimension 2, $0$ mean, and sparse precision matrix:

$$
\pmb{Q} = \begin{bmatrix}
\tau_{\beta_0} & 0\\
0 & \tau_{\beta_1}
\end{bmatrix}
$$ Note that, since $\beta_0$ and $\beta_1$ are fixed effects, the precision parameters $\tau_{\beta_0}$ and $\tau_{\beta_1}$ are fixed.

`r unhide()`
:::

::: callout-note
We can write the linear predictor vector $\pmb{\eta} = (\eta_1,\dots,\eta_N)$ as

$$
\pmb{\eta} = \pmb{A}\pmb{u} = \pmb{A}_1\pmb{u}_1 + \pmb{A}_2\pmb{u}_2 = \begin{bmatrix}
1 \\
1\\
\vdots\\
1
\end{bmatrix} \beta_0 + \begin{bmatrix}
x_1 \\
x_2\\
\vdots\\
x_N
\end{bmatrix} \beta_1
$$

Our linear predictor consists then of two components: an intercept and a slope.
:::


### Simulate example data

We fix the model parameters $\beta_0$, $\beta_1$ and the hyperparameter $\tau_y$ to a given value and simulate the data accordingly using the code below. The simulated response and covariate data are then saved in a `data.frame` object.

```{r}
#| code-fold: show
#| code-summary: "Simulate Data from a LM"

# set seed for reproducibility
set.seed(1234) 

# Fix the model parameters
beta = c(2,0.5)
sd_error = 0.1

# simulate the data
n = 100
x = rnorm(n)
y = beta[1] + beta[2] * x + rnorm(n, sd = sd_error)

# create the data frame object
df = data.frame(y = y, x = x)  

```

### Fitting a linear regression model with `inlabru`

------------------------------------------------------------------------

**Step1: Defining model components**

The first step is to define the two model components: The intercept and the linear covariate effect.

::: {.callout-warning icon="false"}
## {{< bi pencil-square color=#c8793c >}} Task

Define an object called `cmp` that includes and (i) intercept `beta_0` and (ii) a covariate `x` linear effect `beta_1`.

`r hide("Take hint")`

The `cmp` object is here used to define model components. We can give them any useful names we like, in this case, `beta_0` and `beta_1`. You can remove the automatic intercept construction by adding a `-1` in the components

`r unhide()`

```{r}
#| webex.hide: "Click here to see the solution"
#| code-fold: show
#| purl: false
cmp =  ~ -1 + beta_0(1) + beta_1(x, model = "linear")

```
:::

::: callout-note
Note that we have excluded the default Intercept term in the model by typing `-1` in the model components. However, `inlabru` has automatic intercept that can be called by typing `Intercept()` , which is one of `inlabru` special names and it is used to define a global intercept, e.g.

```{r}
#| eval: false
#| purl: false

cmp =  ~  Intercept(1) + beta_1(x, model = "linear")

```
:::


**Step 2: Build the observation model**

The next step is to construct the observation model by defining the model likelihood. The most important inputs here are the `formula`, the `family` and the `data`.

::: {.callout-warning icon="false"}
## {{< bi pencil-square color=#c8793c >}} Task

Define a linear predictor `eta` using the component labels you have defined on the previous task.

`r hide("Take hint")`

The `eta` object defines how the components should be combined in order to define the model predictor.

`r unhide()`

```{r}
#| webex.hide: "Click here to see the solution"
#| code-fold: show
#| purl: false
eta = y ~ beta_0 + beta_1

```
:::

The likelihood for the observational model is defined using the `bru_obs()` function.

::: {.callout-warning icon="false"}
## {{< bi pencil-square color=#c8793c >}} Task

Define the observational model likelihood in an object called `lik` using the `bru_obs()` function.

`r hide("Take hint")`

The `bru_obs` is expecting three arguments:

-   The linear predictor `eta` we defined in the previous task
-   The data likelihood (this can be specified by setting `family = "gaussian"`)
-   The data set `df`

`r unhide()`

```{r}
#| webex.hide: "Click here to see the solution"
#| code-fold: show
#| purl: false
lik =  bru_obs(formula = eta,
            family = "gaussian",
            data = df)

```
:::


**Step 3: Fit the model**

We fit the model using the `bru()` functions which takes as input the components and the observation model:

```{r}
#| code-summary: "Fit LM in `inlabru`"
fit.lm = bru(cmp, lik)
```

**Step 5: Extract results**


There are several ways to extract and examine the results of a fitted `inlabru` object. 

The most natural place to start is to use the 
 `summary()` which gives access to some basic information about model fit and estimates

```{r }
#| code-summary: "Model summaries"
#| collapse: true
summary(fit.lm)
```

We can see that both the intercept and slope and the error precision are correctly estimated.


Another way, which gives access to more complicated (and useful) output is to use the `predict()` function. 

Below we  take the fitted `bru` object and use the `predict()` function to produce predictions for $\mu$ given a new set of values for the model covariates or the original values used for the model fit

```{r}
new_data = data.frame(x = c(df$x, runif(10)),
                      y = c(df$y, rep(NA,10)))
pred = predict(fit.lm, new_data, ~ beta_0 + beta_1,
               n.samples = 1000)
```

The `predict()` function generate samples from the fitted model. In this case we set the number of samples to 1000.

::: panel-tabset
## Plot

```{r}
#| code-fold: true
#| fig-cap: Data and 95% credible intervals
#| echo: false
#| message: false
#| warning: false
#| fig-align: center
#| fig-width: 4
#| fig-height: 4

pred %>% ggplot() + 
  geom_point(aes(x,y), alpha = 0.3) +
  geom_line(aes(x,mean)) +
  geom_line(aes(x, q0.025), linetype = "dashed")+
  geom_line(aes(x, q0.975), linetype = "dashed")+
  xlab("Covariate") + ylab("Observations")
```

## R Code

```{r}
#| code-fold: show
#| eval: false
#| purl: false


pred %>% ggplot() + 
  geom_point(aes(x,y), alpha = 0.3) +
  geom_line(aes(x,mean)) +
  geom_line(aes(x, q0.025), linetype = "dashed")+
  geom_line(aes(x, q0.975), linetype = "dashed")+
  xlab("Covariate") + ylab("Observations")
```
:::

::: {.callout-warning icon="false"}
## {{< bi pencil-square color=#c8793c >}} Task

Generate predictions for the linear predictor $\mu$ when the covariate has value $x_0 = 0.45$.

What is the predicted value for $\mu$? And what is the uncertainty?

`r hide("Take hint")`

You can create a new data frame containing the new observation $x_0$ and then use the `predict` function.

`r unhide()`

```{r}
#| webex.hide: "Click here to see the solution"
#| code-fold: show
#| purl: false

new_data = data.frame(x = 0.45)
pred = predict(fit.lm, new_data, ~ beta_0 + beta_1,
               n.samples = 1000)

pred

```

You can see the predicted mean and sd by examining the produced `pred` object. In this case the mean is ca 2.22 with sd ca 0.01. This gives a 95% CI ca [2.20,  2.25].
:::


**NOTE** Now we have produced a credible interval for the expected mean $\mu$ if we want to produce a _prediction_ interval for  a new observation $y$ we need to add the uncertainty that comes from the likelihood with precision $\tau_y$. To do this we can again use the `predict()` function to compute  a 95% prediction interval for $y$.

```{r}

pred2 = predict(fit.lm, new_data, 
               formula = ~ {
                 mu = beta_0 + beta_1
                 sigma = sqrt(1/Precision_for_the_Gaussian_observations)
                 list(q1 = qnorm(0.025, mean = mu, sd = sigma),
                      q2 =  qnorm(0.975, mean = mu, sd = sigma))},
               n.samples = 1000)
round(c(pred2$q1$mean, pred2$q2$mean),2)
```
Notice that now the interval we obtain is larger.

## Linear model with discrete variables and  interactions{#LM_int}


We consider now the dataset `iris`. Here data are recorded about 150 different iris flowers belonging to 3 different species (50 for each specie).

You can get more information about these data using 
```{r}
#| eval: false
?iris
```

We want to model the `Petal.length` as a function of `Sepal.length` and species.

```{r}
data("iris")

iris %>% ggplot() + geom_point(aes(Sepal.Length, Petal.Length, color= Species)) +
  facet_wrap(.~Species)
```

**Model 1 - Only `Species` effect**
Our first model assumes that the Sepal length only depends on the species, which is a categoriacal variable.

$$
\begin{aligned}
Y_i & \sim\mathcal{N}(\mu_i,\sigma_y),\ &  i = 1,\dots,150 \\
\mu_{i} & = \eta_{i} = \beta_1\ I(\text{obs }i\text{belongs to species } 1  ) + \beta_2\ I(\text{obs }i\text{belongs to species } 2  ) + \beta_3\ I(\text{obs }i\text{belongs to species } 3  )
\end{aligned}
$$

Using `lm()` we can fit the model as:
```{r}
mod1 = lm(Petal.Length ~ Species, data  = iris)
summary(mod1)
```

Notice that `lm()` uses `setosa` as reference category, the parameter `Speciesversicolor` is then interpreted as the difference between the effect of the  reference species and effect of `versicolor` species.

::: {.callout-warning icon="false"}
## {{< bi pencil-square color=#c8793c >}} Task

Implement the model above using `inlabru` using the `model.matrix()` function in R.


```{r}
#| webex.hide: "Click here to see the solution"
#| code-fold: show
#| purl: false

#Option 1 - Use the model.matrix() function

mm = model.matrix(Petal.Length ~ Species, data  = iris)
iris1 = cbind(iris, mm)
cmp = ~ Intercept(1) +  versicolor(Speciesversicolor, model = "linear") + virginica(Speciesvirginica, model  = "linear")
lik = bru_obs(formula =Petal.Length ~ .,
              data = iris1)
fit1a = bru(cmp, lik)


```

:::

Another way to fit this model is to realize that a fixed effect can be seen as an iid effect with fixed precision, one of  the `inlabru` ways to fit the model is:
```{r}

cmp = ~ -1 +  cov(Species, model = "iid", fixed = T, initial = log(0.001))
lik = bru_obs(formula =Petal.Length ~ .,
              data = iris)
fit1b = bru(cmp, lik)
```
Notice that we fix the precision of the `iid` effect to the same value as the precision of the linear effects which is 0.001.




The fitted values can be inspected as

```{r}
fit1b$summary.random$cov
```



The results from the `fit1b` model is not the same we get from the `lm()` fit. What is happening? It is just a matter of parametrization.

In the following we are going to recover the `lm()` results using the fitted `fit1b` object and the `predict()` function.

```{r}
#| webex.hide: "Click here to see the solution"
#| code-fold: show
#| purl: false

newdata = data.frame(Species = 0)
preds = predict(fit1b, newdata = c() , ~ data.frame(Reference = cov_latent[2],
                                             Speciesversicolor = cov_latent[2]-cov_latent[1],
                                             Speciesvirginica = cov_latent[3]-cov_latent[1]))
```


Two things to notice here:

1. We have used a empty object as `newdata` in the `predict()` function.
2. The suffix `_latent` indicates that we are interested in the latent model efffect.

**Model 2 - Interaction between `Species` and `Sepal.Length`**

Out second model is defined as
$$
\begin{aligned}
Y_i & \sim\mathcal{N}(\mu_i,\sigma_y),\ &  i = 1,\dots,150 \\
\mu_{i} & = \eta_{i} = \beta_0 +  \beta_1 x_i\ I(\text{obs }i\text{belongs to species } 1  ) + \beta_2x_i\ I(\text{obs }i\text{belongs to species } 2  ) + \beta_3x_i\ I(\text{obs }i\text{belongs to species } 3  )
\end{aligned}
$$

that is, we have a common intercept $\beta_0$ while the linear effect of the Sepal length depends on the Species. Using `lm()` we have:
```{r}
mod2 = lm(Petal.Length ~ Species:Sepal.Length, data = iris)
mod2
```


::: {.callout-warning icon="false"}
## {{< bi pencil-square color=#c8793c >}} Task

Fit the same model in `inlabru` using the `model.matrix()`.


```{r}
#| webex.hide: "Click here to see the solution"
#| code-fold: show
#| purl: false

mm = model.matrix(Petal.Length ~ Species:Sepal.Length, data = iris)

# change the names to make the R friendly (the : makes problems when using them to identify columns)

colnames(mm) = c("Int", "Setosa_Sepal",  "Versicolor_Sepal", "Virginica_Sepal")

iris2 = cbind(iris, mm)

cmp = ~ Intercept(1) + int1(Setosa_Sepal, model = "linear") +
  int2(Versicolor_Sepal, model = "linear") +
  int3(Virginica_Sepal, model = "linear") 
  
lik = bru_obs(formula =Petal.Length ~ .,
              data = iris2)
fit2 = bru(cmp, lik)


```
:::


A second option is to use a _weighted_ iid random effect with fixed precision as:
```{r}

cmp = ~Intercept(1)  + 
  slope(Species, Sepal.Length, model = "iid", fixed = T, initial = log(0.001))

lik = bru_obs(formula =Petal.Length ~ .,
              data = iris)
fit2b = bru(cmp, lik)
```

Notice that this time, the parametrization of `lm()` and `inlabru` is the same, so we get the same results from the two models.



::: {.callout-warning icon="false"}
## {{< bi pencil-square color=#c8793c >}} Task

Plot the estimated regression lines for the three species using model `fit2b`


```{r}
#| webex.hide: "Click here to see the solution"
#| code-fold: show
#| purl: false

preds = predict(fit2b, iris, ~ Intercept + slope)

pp = preds %>% ggplot() + geom_line(aes(Sepal.Length, mean, group = Species, color = Species)) +
  geom_ribbon(aes(Sepal.Length, ymin = q0.025, ymax = q0.975, 
                  group = Species, fill = Species), alpha = 0.5) +
  geom_point(aes(Sepal.Length, Petal.Length,color = Species))

```
:::


## Linear Mixed Model{#LMM}

Finally we are going to consider a simple linear mixed model, that is 
 a simple linear regression model except with the addition that the data that comes in groups. 
 
 Suppose that we want to include a random effect for each group $j$ (equivalent to adding a group random intercept). The model is then: 

$$
 y_{ij}  = \beta_0 + \beta_1 x_i + u_j + \epsilon_{ij} ~~~  \text{for}~i = 1,\ldots,N~ \text{and}~ j = 1,\ldots,m.
$$

Here the random group effect is given by the variable $u_j \sim \mathcal{N}(0, \tau^{-1}_u)$ with $\tau_u = 1/\sigma^2_u$ describing the variability between groups (i.e., how much the group means differ from the overall mean). Then, $\epsilon_j \sim \mathcal{N}(0, \tau^{-1}_\epsilon)$ denotes the residuals of the model and $\tau_\epsilon = 1/\sigma^2_\epsilon$ captures how much individual observations deviate from their group mean (i.e., variability within group).

The model design matrix for the random effect has one row for each observation (this is equivalent to a random intercept model). The row of the design matrix associated with the $ij$-th observation consists of zeros except for the element associated with $u_j$, which has a one.

$$
\pmb{\eta} = \pmb{A}\pmb{u} = \pmb{A}_1\pmb{u}_1 + \pmb{A}_2\pmb{u}_2 + \pmb{A}_3\pmb{u}_3
$$

::: {.callout-note icon="false"}
## Supplementary material: LMM as a LGM

In matrix form, the linear mixed model for the *j*-th group can be written as:

$$ \overbrace{\mathbf{y}_j}^{ N \times 1} = \overbrace{X_j}^{ N \times 2} \underbrace{\beta}_{1\times 1} + \overbrace{Z_j}^{n_j \times 1} \underbrace{u_j}_{1\times1} + \overbrace{\epsilon_j}^{n_j \times 1}, $$

In a latent Gaussian model (LGM) formulation the mixed model predictor for the *i*-th observation can be written as :

$$
\eta_i = \beta_0 + \beta_1 x_i + \sum_k^K f_k(u_j)
$$

where $f_k(u_j) = u_j$ since there’s only one random effect per group (i.e., a random intercept for group $j$). The fixed effects $(\beta_0,\beta_1)$ are assigned Gaussian priors (e.g., $\beta \sim \mathcal{N}(0,\tau_\beta^{-1})$). The random effects $\mathbf{u} = (u_1,\ldots,u_m)^T$ follow a Gaussian density $\mathcal{N}(0,\mathbf{Q}_u^{-1})$ where $\mathbf{Q}_u = \tau_u\mathbf{I}_m$ is the precision matrix for the random intercepts. Then, the components for the LGM are the following:

-   Latent field given by

    $$
    \begin{bmatrix} \beta \\\mathbf{u}
    \end{bmatrix} \sim \mathcal{N}\left(\mathbf{0},\begin{bmatrix}\tau_\beta^{-1}\mathbf{I}_2&\mathbf{0}\\\mathbf{0} &\tau_u^{-1}\mathbf{I}_m\end{bmatrix}\right)
    $$

-   Likelihood:

    $$
    y_i \sim \mathcal{N}(\eta_i,\tau_{\epsilon}^{-1})
    $$

-   Hyperparameters:

    -   $\tau_u\sim\mathrm{Gamma}(a,b)$
    -   $\tau_\epsilon \sim \mathrm{Gamma}(c,d)$
:::

### **Simulate example data**

```{r}
#| message: false
#| warning: false
#| purl: false
#| echo: false
library(inlabru)
library(INLA)
library(ggplot2)
library(dplyr)
```

```{r }
#| code-summary: "Simulate data from a LMM"
#|
set.seed(12)
beta = c(1.5,1)
sd_error = 1
tau_group = 1

n = 100
n.groups = 5
x = rnorm(n)
v = rnorm(n.groups, sd = tau_group^{-1/2})
y = beta[1] + beta[2] * x + rnorm(n, sd = sd_error) +
  rep(v, each = 20)

df = data.frame(y = y, x = x, j = rep(1:5, each = 20))
```

Note that `inlabru` expects an integer indexing variable to label the groups.

```{r}
#| code-fold: true
#| fig-cap: Data for the linear mixed model example with 5 groups
#| fig-align: center
#| fig-width: 4
#| fig-height: 4

ggplot(df) +
  geom_point(aes(x = x, colour = factor(j), y = y)) +
  theme_classic() +
  scale_colour_discrete("Group")

```




### Fitting a LMM in `inlabru`

------------------------------------------------------------------------

This is done in three steps:

1. Define the model components
2. Define the formula and the observation model (likelihood) using `bru_obs()`
3. Run the model using `bru()`

After that you can collect and inspect results.

**Defining model components and observational model**

In order to specify this model we must use the `group` argument to tell `inlabru` which variable indexes the groups. The `model = "iid"` tells INLA that the groups are independent from one another.

```{r }
# Define model components
cmp =  ~ -1 + beta_0(1) + beta_1(x, model = "linear") +
  u(j, model = "iid")
```

The group variable is indexed by column `j` in the dataset. We have chosen to name this component `v()` to connect with the mathematical notation that we used above.

```{r }
# Construct likelihood
lik =  bru_obs(formula = y ~.,
            family = "gaussian",
            data = df)
```

**Fitting the model**

The model can be fitted exactly as in the previous examples by using the `bru` function with the components and likelihood objects.

```{r}
#| collapse: true
#| code-summary: "Fit a LMM in inlabru"
fit = bru(cmp, lik)
summary(fit)
```

### Model predictions

To compute model predictions we can create a `data.frame` containing a range of values of covariate where we want the response to be predicted for each group. Then we simply call the predict function while specifying the model components.

```{r}
#| code-fold: true
#| code-summary: "LMM fitted values"
#| fig-align: center
#| fig-width: 4
#| fig-height: 4

# New data
xpred = seq(range(x)[1], range(x)[2], length.out = 100)
j = 1:n.groups
pred_data = expand.grid(x = xpred, j = j)
pred = predict(fit, pred_data, formula = ~ beta_0 + beta_1 + u)


pred %>%
  ggplot(aes(x=x,y=mean,color=factor(j)))+
  geom_line()+
  geom_ribbon(aes(x,ymin = q0.025, ymax= q0.975,fill=factor(j)), alpha = 0.5) +
  geom_point(data=df,aes(x=x,y=y,colour=factor(j)))+
  facet_wrap(~j)

```

::: {.callout-tip icon="false"}
## {{< bi question-octagon color=#6dc83c >}} Question

Suppose that we are also interested in including random slopes into our model. Assuming intercept and slopes are independent, can your write down the linear predictor and the components of this model as a LGM?

`r hide("Give me a hint")`

In general, the mixed model predictor can decomposed as:

$$ \pmb{\eta} = X\beta + Z\mathbf{u} $$

Where $X$ is a $n \times p$ design matrix and $\beta$ the corresponding *p*-dimensional vector of fixed effects. Then $Z$ is a $n\times q_J$ design matrix for the $q_J$ random effects and $J$ groups; $\mathbf{v}$ is then a $q_J \times 1$ vector of $q$ random effects for the $J$ groups. In a latent Gaussian model (LGM) formulation this can be written as:

$$ \eta_i = \beta_0 + \sum\beta_j x_{ij} + \sum_k f(k) (u_{ij}) $$

`r unhide()`

`r hide("See Solution")`

-   The linear predictor is given by

    $$
    \eta_i = \beta_0 + \beta_1x_i + u_{0j} + u_{1j}x_i
    $$

-   Latent field defined by:

    -   $\beta \sim \mathcal{N}(0,\tau_\beta^{-1})$

    -   $\mathbf{u}_j = \begin{bmatrix}u_{0j} \\ u_{1j}\end{bmatrix}, \mathbf{u}_j \sim \mathcal{N}(\mathbf{0},\mathbf{Q}_u^{-1})$ where the precision matrix is a block-diagonal matrix with entries $\mathbf{Q}_u= \begin{bmatrix}\tau_{u_0} & {0} \\{0} & \tau_{u_1}\end{bmatrix}$

-   The hyperparameters are then:

    -   $\tau_{u_0},\tau_{u_1} \text{and}~\tau_\epsilon$

To fit this model in `inlabru` we can simply modify the model components as follows:

```{r}
#| eval: false
#| purl: false

cmp =  ~ -1 + beta_0(1) + beta_1(x, model = "linear") +
  u0(j, model = "iid") + u1(j,x, model = "iid")
```

`r unhide()`
:::
